Applied Temporal Analysis: A Complete Run of the FraCaS Test Suite

### 2021-02-09 paper2020/paper.tex
fit

For the purpose of this paper, the important feature of GF
syntax is that it aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
%
Consequently, the system presented
here does not aim at textual natural language understanding, but rather
provides a testable, systematic formal semantics of temporal phenomena.
Example \ref{ex:nobel} shows an example abstract syntax tree and its
realisation in English.

@@

\paragraph{Montagovian Semantics}
Non-withstanding special support for anaphora, the core of the translation of syntax trees
to logical form follows a largely standard
montagovian semantics. In brief, sentences are interpreted as
propositions, verbs and noun-phrases as predicates.  We use
type-raising of noun-phrases, to support
quantifiers~\citep{montague_proper_1974}.

@@

\onelingex{A scandinavian won the nobel prize.\label{ex:nobel}
{\small\begin{multline*}
\varid{useCl}\, \varid{past}\, \varid{pPos}\\ (\varid{predVP}\, (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{scandinavian\_CN})\\ (\varid{complSlash}\, (\varid{slashV2a}\, \varid{win\_V2})\\ (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{nobel\_prize\_CN})))
\end{multline*}
}}
%
In our semantics we deal only with present and past tenses (simple and
continuous). Indeed we find that FraCas does not exercise additional
specific tenses. (When a more complicated tense is used, the
additional information is also carried by adverbs or adverbial
phrases, in a more specific way). While we believe that many other
tenses can be captured under the same general framework, we leave a
detailed study to further work.

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in some way. Sometimes the restriction is an equality, as in ``for
  exactly two hours''. In the following example we show the inclusion
  constraint, for ``in 1992''.
%
  \begin{multline*}
  ⟦\text{in 1992, \(s\)}⟧(\anyspan) = \\ ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)
\end{multline*}
In the FraCaS test suite, we normally do not find several
time-modifying adverbs modifying a single verb phrase. Indeed, sentences such
as ``in 1992, in 1991 john wrote a novel'' are infelicitous. This
justifies ignoring the input timespan in the above interpretation --
we are in particular not interested in modelling felicity with our
semantics, only giving an accurate semantics when the input is
felicitous.
\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include ``always'' and ``never''.

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \begin{multline*}
  ⟦\text{always \(s\)}⟧(t_0,t_1) = \\ ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')
\end{multline*}
Note that here we \emph{do} use the input interval, resulting in a correct interpretation for
phrases such as ``In 1994, Itel was always on time.'' .

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurrences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.
%
For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:
%
\begin{multline}
  \constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → \\ (t_1 = t_3) ∧ (t_2 = t_4)
  \label{eq:unicity}
\end{multline}

@@

\paragraph{Statives}
\textit{A contrario}, if \(P\) is \emph{stative}, then we get a time-interval subsumption property:
\begin{multline*}
\constant {subsumption}_P :\\ [t_3,t_4] ⊆ [t_1,t_2] → P (t_1,t_2) → P (t_3,t_4)
\end{multline*}
This principle is used to reason about problem (314), below:

@@

\begin{equation}
  \label{eq:smith-novel-1}
\begin{array}{l}
∀x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\end{equation}
Note here that the scope for the existential is extended beyond the
scope of \textbf{P1}, and its polarity switched (to universal). This extension can follow the account of
\citet{unger_dynamic_2011}, and our implemented analysis of
anaphora\ifanon (citation anonymized)
\else\citep{bernardy_computational_2018,bernardy_wide-coverage_2019}\fi.

@@

\fracasex{280}{
\item	Smith wrote a novel in 1991.
\hyp 	Smith wrote a novel in 1992.
}
Here our interpretation is:

\[\begin{array}{l}
(∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2)) ∧ \\
(∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4)) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, because $x \neq y$, and thus we cannot
deduce equality of the intervals $t_1,t_2$ and $t_3,t_4$. In turn, the
hypothesis cannot be refuted.

@@

The final class of lexemes carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal context in non-trivial
ways. Consider for example ``finish to ...''. The timespan of the
argument of ``finish'' should end within the timespan of the finishing
action:
\begin{multline*}
⟦\text{finish to \(s\)}⟧(t_0,t_1) = \\ ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')
\end{multline*}

@@

\fracasex{232}
{\item	ITEL won more orders than APCOM did.
\item	APCOM won ten orders.
\hyp 	ITEL won at least eleven orders.
}

@@

Further cases to be included in an extended FraCaS future suite
involve examples where the interaction between different tenses needs
to be captured:

\fracasex{extra3}{
	\item	When the phone rang, John had entered the house.  
	\hyp 	John entered the house before the phone rang (YES).
}

>>>>

For the purpose of this paper, the important feature of GF
syntax is that it aims at a balance of sufficient abstraction to provide a
semantically-relevant structure, but at the same time it embeds
sufficiently many syntactic features to be able to reconstruct
natural-language text. That is, the parse trees generally satisfy the
homomorphism requirement of \citet{montague_english_1970,montague_proper_1974}, and
we can focus on the translation of syntactic trees to logical forms.
%
Consequently, the system presented
here does not aim at textual natural language understanding, but rather
provides a testable, systematic formal semantics of temporal phenomena.
Example \ref{ex:nobel} shows an abstract syntax tree and its
realisation in English.

@@

\paragraph{Montagovian Semantics}
Non-withstanding special support for anaphora, the core of the translation of syntax trees
to logical form follows a standard
montagovian semantics. In brief, sentences are interpreted as
propositions, verbs and noun-phrases as predicates.  We use
type-raising of noun-phrases, to support
quantifiers~\citep{montague_proper_1974}.

@@

\onelingex{A scandinavian won the nobel prize.\label{ex:nobel}
  \vspace{-1.5ex}
{\small\begin{multline*}
\varid{useCl}\, \varid{past}\, \varid{pPos}\\ (\varid{predVP}\, (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{scandinavian\_CN})\\ (\varid{complSlash}\, (\varid{slashV2a}\, \varid{win\_V2})\\ (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{nobel\_prize\_CN})))
\end{multline*}
}}
%
In our semantics we deal only with present and past tenses (simple and
continuous). Indeed we find that FraCas does not exercise additional
specific tenses. (When a more complicated tense is used, the
additional information is also carried by adverbs or adverbial
phrases, in a more specific way). While we believe that many other
tenses can be captured under the same general framework, we leave a
detailed study to further work.

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in some way. Sometimes the restriction is an equality, as in ``for
  exactly two hours''. In the following example we show the inclusion
  constraint, for ``in 1992''.
%
  \vspace{-1.5ex}
  \begin{multline*}
  ⟦\text{in 1992, \(s\)}⟧(\anyspan) = \\ ∃t_1,t_2.  [t_1,t_2] ⊆ 1992, ⟦s⟧(t_1,t_2)
\end{multline*}
In the FraCaS test suite, we normally do not find several
time-modifying adverbs modifying a single verb phrase. Indeed, sentences such
as ``in 1992, in 1991 john wrote a novel'' are infelicitous. This
justifies ignoring the input timespan in the above interpretation --
we are in particular not interested in modelling felicity with our
semantics, only giving an accurate semantics when the input is
felicitous.
\item[universally quantifying] A few adverbs introduce intervals via a
  universal quantification (sometimes with a constraint). Examples
  include ``always'' and ``never''.

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \vspace{-1.5ex}
  \begin{multline*}
  ⟦\text{always \(s\)}⟧(t_0,t_1) = \\ ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')
\end{multline*}
Note that here we \emph{do} use the input interval, resulting in a correct interpretation for
phrases such as ``In 1994, Itel was always on time.'' .

@@

To be able to conclude that there is entailment, as the testsuite expects, we have to make sure
that the two occurrences of ``Jones left'' (in \textbf{P1} and \textbf{P2}) refer to the
same time intervals.
%
For this purpose we postulate \emph{unicity of action} for certain time-dependent
propositions:
%
  \vspace{-1.5ex}
\begin{multline}
  \constant{unicity}_P : P (t_1,t_2) → P (t_3,t_4) → \\ (t_1 = t_3) ∧ (t_2 = t_4)
  \label{eq:unicity}
\end{multline}

@@

\paragraph{Statives}
\textit{A contrario}, if \(P\) is \emph{stative}, then we get a time-interval subsumption property:
\vspace{-1.5ex}
\begin{multline*}
\constant {subsumption}_P :\\ [t_3,t_4] ⊆ [t_1,t_2] → P (t_1,t_2) → P (t_3,t_4)
\end{multline*}
This principle is used to reason about problem (314), below:

@@

\vspace{-1.5ex}
\begin{equation}
  \label{eq:smith-novel-1}
\begin{array}{l}
∀x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,x,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\end{equation}
Note here that the scope for the existential is extended beyond the
scope of \textbf{P1}, and its polarity switched (to universal). This extension can follow the account of
\citet{unger_dynamic_2011}, and our implemented analysis of
anaphora\ifanon (citation anonymized)
\else\citep{bernardy_computational_2018,bernardy_wide-coverage_2019}\fi.

@@

\fracasex{280}{
\item	Smith wrote a novel in 1991.
\hyp 	Smith wrote a novel in 1992.
}
Here our interpretation is:
\vspace{2ex}
\[\begin{array}{l}
(∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2)) ∧ \\
(∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4)) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, because $x \neq y$, and thus we cannot
deduce equality of the intervals $t_1,t_2$ and $t_3,t_4$. In turn, the
hypothesis cannot be refuted.

@@

The final class of lexemes carrying a temporal-dependent semantics are
verbs taking a proposition as argument, like ``finish'', ``start'',
etc. These verbs modify the temporal context in non-trivial
ways. Consider for example ``finish to ...''. The timespan of the
argument of ``finish'' should end within the timespan of the finishing
action:
\vspace{-1.5ex}
\begin{multline*}
⟦\text{finish to \(s\)}⟧(t_0,t_1) = \\ ∃(t_0',t_1'). t_1' ∈ [t_0,t_1] ∧ ⟦s⟧(t_0',t_1')
\end{multline*}

@@

\vspace{-1.5ex}
\fracasex{232}
{\item	ITEL won more orders than APCOM did.
\item	APCOM won ten orders.
\hyp 	ITEL won at least eleven orders.
}

@@

% Further cases to be included in an extended FraCaS future suite
% involve examples where the interaction between different tenses needs
% to be captured:
% \fracasex{extra3}{
% 	\item	When the phone rang, John had entered the house.  
% 	\hyp 	John entered the house before the phone rang (YES).
% }
